---
title: "Explore Gapminder using dplyer and ggplot"
output: 
  html_document:
    keep_md: true
---

```{r}
library(tidyverse)
library(gapminder)
```

#### Smell test the data

Let's explore the data frame first:
```{r database_explore}
typeof(gapminder)
class(gapminder)
```

Gapminder has type: list and class: tbl_df i.e. tibble dataframe

```{r}

ncol(gapminder)
nrow(gapminder)

```

So, the given dataframe has 6 columns which means 6 variables and 1704 rows or observation.

Size of a dataframe can also be figured out using dim() which gives dimensions of a dataframe. So, nrow() and ncol() can be used in a code where columns and rows need to be used as a parameter for indexing separately.

```{r}
dim(gapminder)
```

Data type of each variable:

```{r}
sapply(gapminder,typeof)
```

The above output shows that a dataframe can consist of various data types which is the main difference between a dataframe and matrix.

#### Explore individual variable

* Categorical Variable: continent

> For a categorical variable, possible value can be checked using levels().

```{r categorical_variable}
levels(gapminder$continent)
summary(gapminder$continent)
gapminder %>% 
  select(continent,lifeExp) %>% 
  ggplot(aes(continent)) + geom_bar()
```


> Above output tells how many data points are there for each continent which can also be represted using barplot.


* Quantitative Variable

> The range can be checked using range()(which gives the min-max value of the variable in the dataframe)

```{r quantitative_variable}
range(gapminder$lifeExp)
summary(gapminder$lifeExp)
hist(gapminder$lifeExp)
```

> So, mean life expectancy is 59.47 but it varies from 23.60 to 82.60 with major observations around 65-75.


#### Explore various plot types

Let us explore the scatterplot first.

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `gapminder` |
| __aesthetic mapping__ | 'x' and 'y' |
| __geometric object__  | point |
| scale                 | linear |
| statistical transform | none |
| coordinate system     | rectangular |
| facetting             | none |

I have considered two quantitative variables x = lifeExp and y = gdpPercap

Firstly, let us draw basic scatterplot using these two variables.

```{r scatter_plot}
a <- ggplot(gapminder,aes(lifeExp,gdpPercap))
a + geom_point()
```

As we can see from the graph, the plot is exponential due to skewness of the data . This can be changed to linear trend by using scale() in geom_point or as shown below

```{r}
a + geom_point() +
  scale_y_log10()
```

To understand the trend of data, geom_smooth() function can be used:

```{r}
a + geom_point() +
  scale_y_log10() +
  geom_smooth()
```

geom_smooth fits a model("gam" in this case) to the given data in consideration. This model can be changed using method argument. The gray area around the line is point-wise confidence which can be switched off using "se = false".

```{r}
a + geom_point() +
  scale_y_log10() +
  geom_smooth(se=FALSE)
```

We can also color code the graph for different continents:

```{r}
a + geom_point(aes(color = continent)) +
  scale_y_log10()
```

This gives an idea that how the data from different continents is spread.


**Density plot**

```{r}
gapminder %>% 
  filter(continent=="Europe") %>% 
  ggplot(aes(lifeExp)) +
  geom_freqpoly()
```

Frequency polygon graph shows the count of observations with a certain value of a quantitative variable. For example, there are around 44 observation with lifeExp = 70 in Europe database.

**Jitter Plot**

Lets jump back to categorical and quantitative variable we considered before and dive deep into it.

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `gapminder` |
| __aesthetic mapping__ | 'x' and 'y' |
| __geometric object__  | boxplot |
| scale                 | linear |
| statistical transform | 5-number summary |

```{r}
ggplot(gapminder,aes(continent,lifeExp)) +
  geom_jitter(alpha = 0.5)
gapminder %>% 
  group_by(continent) %>%
  summarise(grp.mean = mean(lifeExp))
```

As we can see from the data above, average life expectancy in Oceania is the highest but it has very less number of observations.

Lets explore lifeExp for a single continent further.


```{r}
gapminder %>% 
  filter(continent == "Asia") %>% 
  ggplot(aes(year,lifeExp)) +
    geom_line(aes(color = country),show.legend = F)
```

The above graph shows trend of lifeExp through 1952-2007 for different countries of Asia continent.

We can also filter a few countries to compare their data.

```{r}
gapminder %>%
  filter(country %in% c("India", "Afghanistan", "Pakistan","Nepal")) %>%
  ggplot(aes(year,lifeExp)) +
  geom_line(aes(group=country,color = country)) +
  geom_point()
```

Here, it is clear that Afghanistan has lowest lifeExp. Also, lifeExp in India had a steeper increase from 1952-1977 and then slowed down.

#### Additional Exercise

For the code line given below, analyst did not get the intented result because he used "==" operator instead of %in%. So, the ouput showed Rawanda and Afghanistan data but for alternating years.
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

He should have used the following command:
```{r}
filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
```